All the concepts behind TrueSkill Through Time: the \proglang{R}, \proglang{Python} and \proglang{Julia} packages (Draft)

### 2020-08-31 article.tex
TTT: primera versión de introducción

\documentclass[article]{jss}

\newif\ifen
\newif\ifes
\newcommand{\en}[1]{\ifen#1\fi}
\newcommand{\es}[1]{\ifes#1\fi}
\entrue
\usepackage[utf8]{inputenc}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{All the concepts behind TrueSkill Through Time: the \proglang{R}, \proglang{Python} and \proglang{Julia} packages}
\Plaintitle{All the concepts behind TrueSkill Through Time: the R, Python and Julia packages}
\Shorttitle{TrueSkill Through Time: the \proglang{R}, \proglang{Python} and \proglang{Julia} packages}

@@

%% - \Abstract{} almost as usual
\en{
\Abstract{
 \pkg{ttt} package is a \proglang{Python} implementation of TrueSkill Through Time (TTT), an enhancement of TrueSkill algorithm, widely used by the video game industry for skill estimation.
 TTT is a probabilistic model for general pairwise-comparison.
 A nice feature is the use of a single graphical model for the entire history of events.
 The posterior distribution is computed based on an efficient variationl algorithm, known as expectation propagation.
 As a result, the inference procedure requires only a few linear-time iterations over the data.
 This algorithm outperforms competing estimations in terms of model evidence, and scales to millions of parameters.
 We offer an easy to use interface, compatible with the current \code{trueskill} package for \proglang{Python}.
}
}

\es{
\Abstract{ Traducir
}
}

\Keywords{TrueSkill, ranking, skill, education, sports, games, Bayesian inference, \proglang{Python}, \proglang{R}}
\Plainkeywords{TrueSkill, ranking, skill, education, sports, games,  Bayesian inference, Python, R}

@@

% \en{What are the skills of the people?}
% \es{¿Cu\'ales son las habilidades de las personas?}
% %   
% \en{How difficult are the challenges they face?}
% \es{¿Cu\'an dif\'iciles son los desaf\'ios que enfrentan?}
% %
% \en{How are their learning curves over time?}
% \es{¿C\'omo son sus curvas de aprendizaje en el tiempo?}
% %
% \en{What is the overall pattern and how variable is it?}
% \es{¿Cu\'al es el patr\'on general y cu\'an variable es?}
% %

\en{Human develop complex skills over the course of a lifetime.}
\es{Los humanos desarrollan habilidades complejas a lo largo de su vida.}
%
\en{This is made possible by a special integration of biological, cognitive and social processes~\citep{_Koster2020}.}
\es{Esto es posible gracias a una integración especial de los procesos biológicos, cognitivos y sociales~\citep{Koster2020}.}
%
\en{Our brains develop first followed by the body because this sequence allows a longer period of learning~\citep{GonzalezForero2018}.}
\es{Nuetros cerebros se desarrollan primero, seguidos por el cuerpo porque esta secuencia permite un período más largo de aprendizaje~\citep{GonzalezForero2018}.}
%
\en{Human life histories are distinguished by long periods of juvenile dependency, short intervals between births, and long postreproductive life span~\citep{Jones2011}.}
\es{Las historias de vida humanas se distinguen por largos períodos de dependencia juvenil, cortos intervalos entre nacimientos y una larga vida posreproductiva~\citep{Jones2011}.}
%
\en{A special cognitive ability allows humans to learn things from others, and transmit innovations through generations.~\citep{Herrmann2007}.}
\es{Una capacidad cognitiva especial le permite a los humanos aprender de los demás y transmitir las innovación a través de la generaciones~\citep{Herrmann2007}.}
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{derex2020-populationStructure}.}
\es{Al ser un proceso poblacional, la adaptación humana también se ve afectada por características demográficas, como el tamaño y estructura de las poblaciones~\citep{derex2020-populationStructure}.}

@@

\en{Knowing how people's abilities change over time is crucial in many areas.}
\es{Conocer cómo cambian la habilidades de la personas en el tiempo es crucial en muchas \'areas.}
%
\en{In educational contexts, accurate ratings can serve to avoid exposing agents to imbalanced challenges, stimulating interest and engagement.}
\es{En contextos educativos, calificaciones precisas pueden servir para evitar exponer a los agentes a desaf\'ios desproporcionados, estimulando as\'i el inter\'es y la participaci\'on.}
%
\en{For research purposes, they can help to refine and test models of human learning.}
\es{Con fines de investigación, ellas pueden ayudar a refinar y testear modelos de aprendizaje humano.}
%
\en{In practice, since skill is a hidden variable, the best we can do is estimate it from its direct observable consequences.}
\es{En la pr\'actica, dado que la habilidad es una variables oculta, lo mejor que podemos hacer es estimarla a partir de sus consecuencias observables m\'as directas.}
%
\en{The most widely used sources of information in education and sports are the outcomes of problem solving and competition between opponents, respectively.}
\es{Las fuentes de informaci\'on m\'as utilizadas en educaci\'on y deportes provienen del producto de resoluci\'on de problemas y de competencia entre oponentes, respectivamente.}
%
\en{The naive solution, frequency of positive results, is not a good indicator essentially because it also depends on the difficulty of the challenges.}
\es{La soluci\'on ingenua, frecuencia de resultados positivos, no es un buen indicador fundamentalmente porque depende tambi\'en de la dificultad de los desaf\'ios.}
%
\en{For that reason, all solutions rely on pairwise comparisons.}
\es{Por esa raz\'on, todos las soluciones se basan en comparaciones entre pares.}
%
\en{Since the first generative models, proposed almost a century ago by~\cite{thurstone1927-comparativeJugement} and~\cite{zermelo1929-translation}, it is assumed that the probability of an observed result, $r$, depends on the difference in performance, $d$, between the agent and their target, $P(\, r_{ij} \,|\, d_{ij} \,)$.}
\es{Ya los primeros modelos generativos, propuestos hace casi un siglo por~\cite{thurstone1927-comparativeJugement} y~\cite{zermelo1929-translation}, suponen que la probabilidad de un resultado observado, $r$, depende de la diferencia de rendimiento, $d$, entre el agente y su objetivo, $P(\, r_{ij} \,|\, d_{ij} \,)$.}
%
\en{The field revitalized 20 years later by the publications of~\cite{bradley1952-pairedComparisions} and~\cite{mosteller1951-remarks}.}
\es{El campo revitaliz\'o 20 a\~nos despu\'es gracias a las publicaciones de \cite{bradley1952-pairedComparisions} y~\cite{mosteller1951-remarks}.}

@@

\en{Nowadays, the most widely used skill estimators are those proposed by~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, and some of the Item-Response Theory variants compiled by~\cite{vanDerLinden2016-handbookIRT_1} and~\cite{fox2010-bayesianIRT}.}
\es{Actualmente, los estimadores de habilidades m\'as utilizados son los propuestos por~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, y algunas de las variantes de la\emph{Item-Response Theory} recopiladas por~\cite{vanDerLinden2016-handbookIRT_1} y~\cite{fox2010-bayesianIRT}.}
%
\en{Although they sometimes choose alternative bell-shape distributions to define the cumulative probability $P(\, d_{ij} > 0 \,)$, the underlying generative model is the same.}
\es{Por más que a veces se usen distintas distribuciones \emph{bell-shape} para definir la probabilidad acumulada $P(\, d_{ij} > 0 \,)$, el modelo generativo subyacente es el mismo.}
%
\en{The main difference betweem them are methodological: the procedure to solve the shared probabilistic model.}
\es{Las principales diferencias ente ellos son metodológicas: el procedimiento para resolver el modelo probabilístico compartido.}
% Cambio de parrafo.
\begin{figure}[h!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r_{ij}$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r_{ij} = (d_{ij}>0)$};

@@

\en{The causal relationship is not direct.}
\es{La relaci\'on causal no es directa.}

@@

\en{Although agents may have a specific skill the performance they exhibit in an event depends on an undetermined number of unknown factors, which are modeled with a constant noise, $\N(p\,|\,s,\beta^2)$.}
\es{Si bien los agentes tienen una habilidad, el desempe\~no que exhiben en cada evento depende de una cantidad indeterminada de factores desconocidos, que se modelan con un ruido constante, $\N(p\,|\,s,\beta^2)$.}
\en{The observed result is a direct consequence of the difference of those hidden performances, $d_{ij}=p_i - p_j$.}
\es{El resultado observado es consecuencias directa de la diferencia de esos rendimientos ocultos, $d_{ij}=p_i - p_j$.}

@@

\en{Elo's methodological solution extremely simple and astute: update the current estimate based on our surprise.}

@@

\en{Where $y_{ij}$ is $1$ for winners and $-1$ for losers, and the probability of the observed result given the a priori estimates are derived from Figure~\ref{fig:generative_model} (details at section~\ref{}).}
\es{Donde $y_{ij}$ es $1$ para ganadores y $-1$ para perdedores y la probabilidad del resultado observado dadas las estimaciones a priori se deriva de la figura~\ref{fig:generative_model} (detalles en la sección~\ref{}).}

@@

\en{An ad-hoc solution was proposed to break the symmetry of equation~\ref{eq:elo_update}: reduce the impact of $\Delta$ based on how many times the agent was already seen by the system.}
\es{Una soluci\'on ad-hoc fue propuesta para romper la simetr\'ia de la ecuaci\'on~\ref{eq:elo_update}: reducir el impacto de $\Delta$ en funci\'on de las veces que el agente ya fue visto por el sistema.}

@@

% \en{Probability Theory provides us with a consistent framework for quantifying the uncertainty of our beliefs~\cite{bishop2006-PRML}.}
% \es{La Teor\'ia de la Probabilidad nos proporciona un marco consistente para cuantificar la incertidumbre de nuestras creencias~\citep{bishop2006-PRML}.}
% %
\en{The use of Probability Theory to handle uncertainty, known as Bayesian inference, was widely adopted because it has proven to be a extremely useful tool in practice.}
\es{El uso de la Teoría de la Probabilidad para manejar la incertidumbre, conocida como inferencia bayesiana, se adoptó ampliamente porque ha demostrado ser una herramienta extremadamente útil en la práctica.}
%
\en{However, it is not an ad-hoc choice.}
\es{Sin embargo, no una elecci\'on ad-hoc.}
%
\en{Any system that represent degrees of belief by a numerical value, consistent with classical deductive logic and a set of axioms encoding commonsense reasoning, will necessarily be isomorphic to Probability Theory~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
\es{Cualquier sistema que represente grados de creencia mediante un valor num\'erico, consistente con la l\'ogica deductiva cl\'asica y un conjunto de axiomas que codifican el razonamiento de sentido común,  ser\'a necesariamente isom\'orfico a la Teor\'ia de la Probabilidad~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
%
\en{All of the probabilistic inference, no matter how complex, can be solved in terms of two simple equations: the~\ref{eq:sum_rule} and the~\ref{eq:product_rule}.}
\es{Toda inferencia probabil\'istica, no importa cuan compleja ella sea, puede ser resuelta a trav\'es de dos simples ecuaciones: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}

@@

 \underbrace{p(\overbrace{\text{\en{Skills}\es{Habilidades}}}^{\text{\en{Hidden}\es{Oculta}}}|\overbrace{\text{Result\es{ado}}}^{\text{Observ\en{ed}\es{ado}}}, \text{Model\es{o}})}_{\text{Posterior}} = \frac{\overbrace{P(\,\text{Result\es{ado}}\,|\,\text{\en{Skills}\es{Habilidades}}\,,\text{Model\es{o}})}^{\text{\en{Likelihood}\es{Verosimilitud}}}\overbrace{p(\text{\en{Skills}\es{Habilidades}})}^{\text{Prior}}}{\underbrace{P(\text{Result\es{ado}}\,|\,\text{Model\es{o}})}_{\text{Evidenc\en{e}\es{ia} o\en{r}\es{ predicción a} prior \en{predictive}}}}

@@

\en{In section~\ref{} we will see in detail how equation~\ref{eq:inference} is solved in general.}
\es{En la sección~\ref{} veremos en detalle cómo se resuelve la ecuación~\ref{eq:inference} en general.}

@@

\en{The mass that is assigned a posteriori to each hypothesis $s_i$ is proportional to the product of the prior and the likelihood.}
\es{La masa que se le asigna a posteriori a cada hipótesis $s_i$ es proporcional al producto del priori y la verosimilitud.}

@@

\en{Therefore, the likelihood compute all the mass up to the value 0 of a Gaussian distribution centered on the difference between the skill hypothesis and the opponent's mean estimate, ($\delta(s_i) = s_i - \mu_j$), with a variance that includes all uncertainties except the one of $s_i$, ($\vartheta_i^2$).}

@@

\en{The posterior receives the mass of the prior no filtered by the likelihood.}

@@

\en{For each hypothesis $s_i$, the filtered mass is exactly proportional to the magnitude of the surprise, defined as the complement of the likelihood.}

@@

\en{At the region of very high skill hypotheses, where the winning result would have generated almost no surprise, the posterior receives almost all the mass of the prior.}

@@

\en{Conversely, at the region of very low skill hypotheses, where the winning result would have generated a lot of surprise, the posterior receives almost no mass from the prior.}

@@

\en{It is important to stress that the posterior, although similar, is not a Gaussian distribution.}
\es{Es importante remarcar que la posterior, aunque se parezca, no es una distribución Gaussiana.}
%
\en{And without a Gaussian distribution we are prevented from using the equation~\ref{eq:posterior_win} iteratively.}
\es{Y sin distribuciones Gaussianas nos vemos impedidos de usar la ecuación~\ref{eq:posterior_win} iterativamente.}
%
\en{This problem can be addressed through sampling techniques, as~\cite{fox2010-bayesianIRT} does to solve the Bayesian Item-Response Theory (BIRT) variants.}
\es{Este problema puede ser tratado mediante técnicas de sampleo, tal como hace~\cite{fox2010-bayesianIRT} para resolver las variantes Bayesianas de Item-Response Theory (BIRT).}
%
\en{Although sampling methodologies have the property of converging to the exact distribution, they demand a high computational cost.}
\es{Si bien las metodologías de sampleo tienen la propiedad de converger a la distribución exacta, exigen un alto costo computacional.}
%
\en{Since the shape of the exact posterior can be approximated well enough with a Gaussian, the computational cost can be avoided by implementing deterministic approximation techniques.}
\es{Dado que la forma del posterior exacto puede ser aproximado suficientemente bien con una Gaussiana, el costo computacional puede ser evitado implementando técnicas de aproximación determinística.}
%
\en{The main contribution of the Glicko system~\cite{glikman_gliko_2} was the development of an efficient method to approximate the exact posterior with a Gasussian distribution.}
\es{El principal aporte del sistema Glicko~\citep{glikman_gliko_2} fue el desarrolló de un método eficiente para aproximar la posterior exacta con una distribución Gasussiana.}
%
\en{However, the methodology offers no guarantee that the selected Gaussian is the one that best approximates the exact posterior.}
\es{Sin embargo, la metodología no ofrece garantías de que la Gaussiana seleccionada sea la que mejor aproxima al posterior exacto.}
%
\en{The success of the TrueSkill solution~\citep{herbrich2006-trueskill} is based on the application of an efficient method for calculating the Gaussian that best approximates the exact posterior.}
\es{El éxito de la solución TrueSkill~\citep{herbrich2006-trueskill} se basa en la aplicación de método eficiente para calcular la Gaussiana que mejor aproxima a la posterior exacta.}
%
\en{Methodological details are described in the section~\ref{}.}
\es{Los detalles metodológicos se describen en la sección~\ref{}.}
%
\en{This technical advance allowed to apply the equation~\ref{eq:posterior_win} iteratively over a sequence of observations.}
\es{Este avance técnico permitió aplicar la ecuación~\ref{eq:posterior_win} sobre una secuencia de observaciones.}

@@

\en{In this way, information from the past propagates to future estimates.}
\es{De esta forma, la información del pasado propaga hacia las estimaciones futuras.}

@@

\en{However, the imperfect representation of the temporal problem, as a sequence of independent events, introduces a series of problems.}
\es{Sin embargo, la representación imperfecta del problema temporal, como secuencia de eventos independientes, introduce una serie de problemas.}

@@

\en{All weaknesses arise from the fact that in the "filtering approach" the information propagates in the system in only one direction.}
\es{Todas las debilidades surgen por el hecho de que en el "enfoque de filtrado" la información propaga en el sistema en un solo sentido.}

@@

\en{The most obvious problem is that the beginning of any estimation sequence always has high uncertainty, which makes them useless.}
\es{El problema más evidente es que el inicio de toda secuencia de estimaciones siempre tiene alta incertidumbre, lo que las hace inservibles.}

@@

\en{There are, however, a number of less obvious problems to keep in mind. }
\es{Existe sin embago una serie de problemas menos evidentes muy importantes a tener en mente.}

@@

\en{In general terms, filtering approach estimates can suffer from ``temporary'' and ``spatial'' decoupling.}
\en{En términos generales, las estimaciones del enfoque de filtrado pueden sufrir desacoplamientos ``temporales'' y ``espaciales''.}

@@

\en{To understand them it is important to remember that the observed results depend only on the difference in skills, and not on their absolute value.}

>>>>

\documentclass[article]{jss}
% \usepackage{draftwatermark}
% \SetWatermarkText{Draft}
% \SetWatermarkScale{2}
\newif\ifen
\newif\ifes
\newcommand{\en}[1]{\ifen#1\fi}
\newcommand{\es}[1]{\ifes#1\fi}
\entrue
\usepackage[utf8]{inputenc}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{All the concepts behind TrueSkill Through Time: the \proglang{R}, \proglang{Python} and \proglang{Julia} packages (Draft)}
\Plaintitle{All the concepts behind TrueSkill Through Time: the R, Python and Julia packages (Draft)}
\Shorttitle{TrueSkill Through Time: the \proglang{R}, \proglang{Python} and \proglang{Julia} packages (Draft)}

@@

%% - \Abstract{} almost as usual
\Abstract{
 \en{Humans develop complex skills through time.}
 \es{Los humanos desarrollan habilidades complejas a través del tiempo.}
 %
 \en{Knowing how they change is crucial in many areas, such as educational systems and the video game industry.}
 \es{Saber cómo cambian es crucial en muchas áreas, como los sistemas educativos y la industria de los videojuegos.}
 %
 \en{All widely used skill estimators share the same underlying causal model, but differ in their inferential procedure.}
 \es{Todos los estimadores de habilidad ampliamente utilizados comparten el mismo modelo causal subyacente, pero se diferencian en sus procedimientos metodológicas.}
 %
 \en{The success of TrueSkill was based on the application of an efficient algorithm to find the best approximation of the exact posterior.}
 \es{El éxito del TrueSkill se basó en la aplicación de un algoritmo eficiente para encuentrar la mejor aproximación del posterior exacto.}
 %
 \en{TrueSkill Through Time is a major enhancement that uses a single generative model for the entire history of activities, rather than a sequence of independent observations, allowing historical information to propagate throughout the system, resulting in better estimates with even less data.}
 \es{TrueSkill Through Time utiliza un único modelo generativo para toda la historia de las actividades, en lugar de una secuencia de observaciones independientes, lo que permite que la información histórica se propague por todo el sistema, lo que da lugar a mejores estimaciones con aún menos datos.}
 %
 \en{The use of an efficient algorithm, that requires only a few linear iterations over the data, allows scaling to millions of observations in few minutes.}
 \es{El uso de un algoritmo eficiente, que requiere sólo unas pocas iteraciones lineales sobre los datos, permite escalar a millones de observaciones en pocos minutos.}
 %
 \en{This paper offer the first packages for \proglang{R}, \proglang{Python} and \proglang{Julia}, together with its scientific documentation.}
 \es{Mediante este artículo ofrecemos los primeros paquetes para \proglang{R}, \proglang{Python} y \proglang{Julia}, junto con su documentación científica.}
}
\Keywords{Life history, skill, learning, education, sports, gaming, inference, \proglang{R}, \proglang{Python}, \proglang{Julia}}
\Plainkeywords{}

@@

\en{Human develop complex skills over the course of a lifetime.}
\es{Los humanos desarrollan habilidades complejas a lo largo de su vida.}
%
\en{This is made possible by a special integration of biological, cognitive and social processes~\citep{Koster2020}.}
\es{Esto es posible gracias a una integración especial de los procesos biológicos, cognitivos y sociales~\citep{Koster2020}.}
%
\en{Our brains develop first followed by the body because this sequence allows a longer period of learning~\citep{GonzalezForero2018}.}
\es{Nuetros cerebros se desarrollan primero, seguidos por el cuerpo porque esta secuencia permite un período más largo de aprendizaje~\citep{GonzalezForero2018}.}
%
\en{Human life histories are distinguished by long periods of juvenile dependency, short intervals between births, and long postreproductive life span~\citep{Jones2011}.}
\es{Las historias de vida humanas se distinguen por largos períodos de dependencia juvenil, cortos intervalos entre nacimientos y una larga vida posreproductiva~\citep{Jones2011}.}
%
\en{A special cognitive ability allows humans to learn things from others, and transmit innovations through generations.~\citep{Herrmann2007}.}
\es{Una capacidad cognitiva especial le permite a los humanos aprender de los demás y transmitir las innovación a través de la generaciones~\citep{Herrmann2007}.}
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{derex2020-populationStructure}.}
\es{Al ser un proceso poblacional, la adaptación humana también se ve afectada por características demográficas, como el tamaño y estructura de las poblaciones~\citep{derex2020-populationStructure}.}

@@

\en{Knowing how skills change over time is crucial in many areas.}
\es{Conocer cómo cambian la habilidades de la personas en el tiempo es crucial en muchas \'areas.}
%
\en{In educational contexts, accurate ratings can serve to avoid exposing agents to imbalanced challenges, stimulating interest and engagement.}
\es{En contextos educativos, calificaciones precisas pueden servir para evitar exponer a los agentes a desaf\'ios desproporcionados, estimulando as\'i el inter\'es y la participaci\'on.}
%
\en{For research purposes, they can help to refine and test models of human learning.}
\es{Con fines de investigación, ellas pueden ayudar a refinar y testear modelos de aprendizaje humano.}
%
\en{In practice, since skill is a hidden variable, the best we can do is estimate it from its direct observable consequences.}
\es{En la pr\'actica, dado que la habilidad es una variables oculta, lo mejor que podemos hacer es estimarla a partir de sus consecuencias observables m\'as directas.}
%
\en{The most widely used sources of information in education and sports are the outcomes of problem solving and competition between opponents.}
\es{Las fuentes de informaci\'on m\'as utilizadas en educaci\'on y deportes provienen del producto de resoluci\'on de problemas y de competencia entre oponentes.}
%
\en{The naive solution, frequency of positive results, is not a good indicator essentially because it also depends on the difficulty of the challenges.}
\es{La soluci\'on ingenua, frecuencia de resultados positivos, no es un buen indicador fundamentalmente porque depende tambi\'en de la dificultad de los desaf\'ios.}
%
\en{For that reason, all solutions rely on pairwise comparisons.}
\es{Por esa raz\'on, todos las soluciones se basan en comparaciones entre pares.}
%
\en{Since the first generative models, proposed almost a century ago by~\cite{thurstone1927-comparativeJugement} and~\cite{zermelo1929-translation}, it is assumed that the probability of an observed result, $r$, depends on the difference in performance, $d$, between the agent and their target, $P(\, r_{ij} \,|\, d_{ij} \,)$.}
\es{Ya los primeros modelos generativos, propuestos hace casi un siglo por~\cite{thurstone1927-comparativeJugement} y~\cite{zermelo1929-translation}, suponen que la probabilidad de un resultado observado, $r$, depende de la diferencia de rendimiento, $d$, entre el agente y su objetivo, $P(\, r_{ij} \,|\, d_{ij} \,)$.}
%
\en{The field revitalized 20 years later by the publications of~\cite{bradley1952-pairedComparisions} and~\cite{mosteller1951-remarks}.}
\es{El campo revitaliz\'o 20 a\~nos despu\'es gracias a las publicaciones de \cite{bradley1952-pairedComparisions} y~\cite{mosteller1951-remarks}.}

@@

\en{Nowadays, the most widely used skill estimators were proposed by~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, and some of the Item-Response Theory variants compiled by~\cite{vanDerLinden2016-handbookIRT_1} and~\cite{fox2010-bayesianIRT}.}
\es{Actualmente, los estimadores de habilidades m\'as utilizados son los propuestos por~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, y algunas de las variantes de la\emph{Item-Response Theory} recopiladas por~\cite{vanDerLinden2016-handbookIRT_1} y~\cite{fox2010-bayesianIRT}.}
%
\en{Although they sometimes choose alternative bell-shape distributions to define the cumulative probability $P(\, d_{ij} > 0 \,)$, the underlying generative model is the same.}
\es{Por más que a veces se usen distintas distribuciones \emph{bell-shape} para definir la probabilidad acumulada $P(\, d_{ij} > 0 \,)$, el modelo generativo subyacente es el mismo.}
%
\en{The main difference between them are methodological: the procedure to solve the shared probabilistic model.}
\es{Las principales diferencias ente ellos son metodológicas: el procedimiento para resolver el modelo probabilístico compartido.}
% Cambio de parrafo.
\begin{figure}[h!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r_{ij}$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r_{ij} = (d_{ij}>0)$};

@@

% \en{The causal relationship is not direct.}
% \es{La relaci\'on causal no es directa.}
% %
\en{However, the performance depends on an undetermined number of unknown factors, which are modeled with a constant noise, $\N(p\,|\,s,\beta^2)$.}
\es{El desempe\~no depende de una cantidad indeterminada de factores desconocidos, que se modelan con un ruido constante, $\N(p\,|\,s,\beta^2)$.}

@@

\en{Then, the observed result is a direct consequence of the difference of those hidden performances, $d_{ij}=p_i - p_j$.}
\es{Luego, el resultado observado es consecuencias directa de la diferencia de esos rendimientos ocultos, $d_{ij}=p_i - p_j$.}

@@

\en{Elo's methodological solution is extremely simple and astute: update the current estimate based on our surprise.}

@@

\en{Where $y_{ij}$ is $1$ for winners and $-1$ for losers, and the probability of the observed result given the a priori estimates can be derived from Figure~\ref{fig:generative_model} (details at section~\ref{sec:analyticalSolution}).}
\es{Donde $y_{ij}$ es $1$ para ganadores y $-1$ para perdedores y la probabilidad del resultado observado dadas las estimaciones a priori se puede derivar de la figura~\ref{fig:generative_model} (detalles en la sección~\ref{sec:analyticalSolution}).}

@@

\en{to break the symmetry of equation~\ref{eq:elo_update}, nn ad-hoc solution was proposed: reduce the impact of $\Delta$ based on how many times the agent was already seen by the system.}
\es{Para romper la simetr\'ia de la ecuaci\'on~\ref{eq:elo_update}, una soluci\'on ad-hoc fue propuesta: reducir el impacto de $\Delta$ en funci\'on de las veces que el agente ya fue visto por el sistema.}

@@

% \en{Probability Theory provides us with a consistent framework for quantifying the uncertainty of our beliefs~\cite{bishop2006-PRML}.}
% \es{La Teor\'ia de la Probabilidad nos proporciona un marco consistente para cuantificar la incertidumbre de nuestras creencias~\citep{bishop2006-PRML}.}
% %
\en{The use of Probability Theory to handle uncertainty, known as Bayesian inference, was widely adopted because it has proven to be a extremely useful tool in practice.}
\es{El uso de la Teoría de la Probabilidad para manejar la incertidumbre, conocida como inferencia bayesiana, se adoptó ampliamente porque ha demostrado ser una herramienta extremadamente útil en la práctica.}
%
\en{But this is also a principled choice.}
\es{Esto no es una elecci\'on ad-hoc.}
%
\en{Any system that represent degrees of belief by a numerical value, consistent with classical deductive logic and a set of axioms encoding commonsense reasoning, will necessarily be isomorphic to Probability Theory~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
\es{Cualquier sistema que represente grados de creencia mediante un valor num\'erico, consistente con la l\'ogica deductiva cl\'asica y un conjunto de axiomas que codifican el razonamiento de sentido común,  ser\'a necesariamente isom\'orfico a la Teor\'ia de la Probabilidad~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
%
\en{All of the probabilistic inference, no matter how complex, can be solved in terms of two simple equations: the~\ref{eq:sum_rule} and the~\ref{eq:product_rule}.}
\es{Toda inferencia probabil\'istica, no importa cuan compleja ella sea, puede ser resuelta a trav\'es de dos simples ecuaciones: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}

@@

 \underbrace{p(\overbrace{\text{\en{Skills}\es{Habilidades}}}^{\text{\en{Hidden}\es{Oculta}}}|\overbrace{\text{Result\es{ado}}}^{\text{Observ\en{ed}\es{ado}}}, \text{Model\es{o}})}_{\text{Posterior}} = \frac{\overbrace{P(\,\text{Result\es{ado}}\,|\,\text{\en{Skills}\es{Habilidades}}\,,\text{Model\es{o}})}^{\text{\en{Likelihood}\es{Verosimilitud}}}\overbrace{p(\text{\en{Skills}\es{Habilidades}})}^{\text{Prior}}}{\underbrace{P(\text{Result\es{ado}}\,|\,\text{Model\es{o}})}_{\text{Evidenc\en{e}\es{ia} o\en{r}\es{ predicción a} prior \en{prediction}}}}

@@

\en{In section~\ref{sec:analyticalSolution} we will see in detail how equation~\ref{eq:inference} is solved in general.}
\es{En la sección~\ref{sec:analyticalSolution} veremos en detalle cómo se resuelve la ecuación~\ref{eq:inference} en general.}

@@

\en{The density assigned a posteriori to each hypothesis $s_i$ is proportional to the product of the prior and the likelihood.}
\es{La densidad asignada a posteriori a cada hipótesis $s_i$ es proporcional al producto del priori y la verosimilitud.}

@@

\en{Therefore, the likelihood compute all the density up to the value 0 of a Gaussian distribution centered on the difference between the skill hypothesis and the opponent's mean estimate, ($\delta(s_i) = s_i - \mu_j$), with a variance that includes all uncertainties except the one of $s_i$, ($\vartheta_i^2$).}

@@

\en{The posterior receives the density of the prior no filtered by the likelihood.}

@@

\en{For each hypothesis $s_i$, the filtered density is exactly proportional to the magnitude of the surprise, defined as the complement of the likelihood.}

@@

\en{At the region of very high skill hypotheses, where the winning result would have generated almost no surprise, the posterior receives almost all the density of the prior.}

@@

\en{At the region of very low skill hypotheses, where the winning result would have generated a lot of surprise, the posterior receives almost no density from the prior.}

@@

\en{It is important to stress that the posterior, although similar, is not a Gaussian distribution.}
\es{Es importante remarcar que la posterior, aunque se parezca, no es una distribución Gaussiana.}
%
\en{And without a Gaussian distribution we are prevented from using the equation~\ref{eq:posterior_win} iteratively.}
\es{Y sin distribuciones Gaussianas nos vemos impedidos de usar la ecuación~\ref{eq:posterior_win} iterativamente.}
%
\en{This problem can be addressed through sampling techniques, as~\cite{fox2010-bayesianIRT} does to solve the Bayesian Item-Response Theory (BIRT) variants.}
\es{Este problema puede ser tratado mediante técnicas de sampleo, tal como hace~\cite{fox2010-bayesianIRT} para resolver las variantes Bayesianas de Item-Response Theory (BIRT).}
%
%\en{Although sampling methodologies have the property of converging to the exact distribution, they demand a high computational cost.}
%\es{Si bien las metodologías de sampleo tienen la propiedad de converger a la distribución exacta, exigen un alto costo computacional.}
%
\en{Since the shape of the exact posterior can be approximated well enough with a Gaussian, the computational cost of the sampling methodologies can be avoided by implementing deterministic approximation techniques.}
\es{Dado que la forma del posterior exacto puede ser aproximado suficientemente bien con una Gaussiana, el costo computacional de la metodologías de sampleo puede ser evitado implementando técnicas de aproximación determinística.}
%
\en{The main contribution of the Glicko system~\cite{glikman_gliko_2} was the development of an efficient method to approximate the exact posterior with a Gasussian distribution.}
\es{El principal aporte del sistema Glicko~\citep{glikman_gliko_2} fue el desarrolló de un método eficiente para aproximar la posterior exacta con una distribución Gasussiana.}
%
\en{However, the methodology offers no guarantee that the selected Gaussian is the one that best approximates the exact posterior.}
\es{Sin embargo, la metodología no ofrece garantías de que la Gaussiana seleccionada sea la que mejor aproxima al posterior exacto.}
%
\en{The success of the TrueSkill solution~\citep{herbrich2006-trueskill} is based on the application of an efficient method for calculating the Gaussian that best approximates the exact posterior.}
\es{El éxito de la solución TrueSkill~\citep{herbrich2006-trueskill} se basa en la aplicación de método eficiente para calcular la Gaussiana que mejor aproxima a la posterior exacta.}
%
\en{Methodological details are described in the section~\ref{sec:approximateSolution}.}
\es{Los detalles metodológicos se describen en la sección~\ref{sec:approximateSolution}.}
%
\en{This technical advance allowed to apply the equation~\ref{eq:posterior_win} iteratively over a sequence of observations.}
\es{Este avance técnico permitió aplicar la ecuación~\ref{eq:posterior_win} sobre una secuencia de observaciones.}

@@

\en{In this way, information propagates from past to future estimates.}
\es{De esta forma, la información propaga del estimaciones pasadas hacia futuras.}

@@

\en{However, the imperfect representation of the temporal process, as a sequence of independent events, introduces a series of problems.}
\es{Sin embargo, la representación imperfecta del proceso temporal, como secuencia de eventos independientes, introduce una serie de problemas.}

@@

\en{All weaknesses of the filtering approach are due to the fact that information propagates in the system in only one direction.}
\es{Todas las debilidades del enfoque \emph{filtering} se deben a que la información propaga en el sistema en un solo sentido.}

@@

\en{The most obvious consecuence is that the beginning of any sequence of estimations always has high uncertainty, making them useless.}
\es{El consecuencia más evidente es que el inicio de toda secuencia de estimaciones siempre tiene alta incertidumbre, lo que las hace inservibles.}

@@

\en{There are, however, others less obvious problems to keep in mind. }
\es{Existe sin embago otros problemas menos evidentes a tener en mente.}

@@

\en{In general terms, the filtering-approach estimates can suffer from ``temporary'' and ``spatial'' decoupling.}
\es{En términos generales, las estimaciones del enfoque de filtrado pueden sufrir desacoplamientos ``temporales'' y ``espaciales''.}

@@

\en{To understand why it is important to remember that the observed results depend only on the difference in skills, and not on their absolute value.}
